Grove: A Bidirectionally Typed Collaborative Structure Editor Calculus

### 2024-11-22 sections/abstract.tex
directory restructuring and hiding appendices

\begin{abstract}
Version control systems typically rely on a \emph{patch language}, heuristic \emph{patch synthesis algorithms} like \li{diff}, and \emph{three-way merge algorithms}. Standard patch languages and merge algorithms often fail to identify conflicts correctly when there are multiple edits to one line of code or code is relocated. This paper introduces Grove, a collaborative structure editor calculus  that eliminates patch synthesis and three-way merge algorithms entirely. 
Instead, patches are derived directly from the log of the developer's edit actions 
and all edits commute, i.e. the repository state forms a commutative replicated data type (CmRDT). To handle conflicts that can arise due to code relocation, 
the core datatype in Grove is a labeled directed multi-graph with uniquely identified 
vertices and edges. All edits amount to edge insertion and deletion, with deletion being permanent. 
To support tree-based editing, we define a decomposition from graphs into \emph{groves}, which are a set of syntax trees with conflicts---including local, relocation, and unicyclic relocation conflicts--- represented explicitly using holes and references between trees. Finally, we define a type error localization system for groves that enjoys a \emph{totality} property, i.e. all editor states in Grove are statically meaningful, so developers can use standard editor services while 
working to resolve these explicitly represented conflicts. The static semantics is defined as a bidirectional marking system in line with recent work, with gradual typing employed to handle situations where errors and conflicts 
prevent type determination. We then layer on a unification-based type inference system  to opportunistically fill type holes and fail gracefully when no solution exists. We mechanize the metatheory of Grove using the Agda theorem prover. We  implement these ideas as the \emph{Grove Workbench}, which generates the necessary data structures and algorithms in OCaml given a syntax tree specification.
\end{abstract}

>>>>

\begin{abstract}
Version control systems typically rely on a \emph{patch language}, heuristic \emph{patch synthesis algorithms} like \li{diff}, and \emph{three-way merge algorithms}. Standard patch languages and merge algorithms often fail to identify conflicts correctly when there are multiple edits to one line of code or code is relocated. This paper introduces Grove, a collaborative structure editor calculus  that eliminates patch synthesis and three-way merge algorithms entirely. 
Instead, patches are derived directly from the log of the developer's edit actions 
and all edits commute, i.e. the repository state forms a commutative replicated data type (CmRDT). To handle conflicts that can arise due to code relocation, 
the core datatype in Grove is a labeled directed multi-graph with uniquely identified 
vertices and edges. All edits amount to edge insertion and deletion, with deletion being permanent. 
To support tree-based editing, we define a decomposition from graphs into \emph{groves}, which are a set of syntax trees with conflicts---including local, relocation, and unicyclic relocation conflicts---represented explicitly using holes and references between trees. Finally, we define a type error localization system for groves that enjoys a \emph{totality} property, i.e. all editor states in Grove are statically meaningful, so developers can use standard editor services while 
working to resolve these explicitly represented conflicts. The static semantics is defined as a bidirectional marking system in line with recent work, with gradual typing employed to handle situations where errors and conflicts 
prevent type determination. We then layer on a unification-based type inference system  to opportunistically fill type holes and fail gracefully when no solution exists. We mechanize the metatheory of Grove using the Agda theorem prover. We  implement these ideas as the \emph{Grove Workbench}, which generates the necessary data structures and algorithms in OCaml given a syntax tree specification.
\end{abstract}

### 2024-11-22 sections/formalism.tex
directory restructuring and hiding appendices

We now formally consider the ideas presented informally above by defining the Grove Calculus. The calculus consists of a commutative patch language on graphs, defined in \autoref{sub:Convergent Graphs}, a decomposition procedure to go from graphs to groves (i.e. sets of terms with conflicts) in \autoref{sub:Groves}, and a user action language that operates on groves and is defined by translation to the patch language in \autoref{sub:User Actions}.

@@

@@

A \emph{location} $\ell = (v, p) \in \L$, represents the origin of an edge, where $v=(u,k)$ is drawn from $\V$ and position $p$ is drawn from $\mathsf{arity}(k)$.

An \emph{edge} $\e{=}(w, \ell, v) \in \E = \W \times \L \times \V$ represents a directed multi-edge 
    identified by $w$, 
    originating from location $\ell$, 
    with destination vertex $v$.

A \emph{graph} $g : \E \rightarrow \Sigma$ is a function from edges to \emph{edge states} $\Sigma = \SetOf{\bot, \Plus, \Minus}$. If $g(\e) = \bot$, then $\e$ has not yet been constructed in the graph. If $g(\e) = \Plus$ then $\e$ is \textit{live} in the graph, meaning it is relevant to the user and the type system. If $g(\e) = \Minus$, then $\e$ has been deleted and cannot be constructed again. The total ordering $\bot \sqsubset \Plus \sqsubset \Minus$ forms a lattice over $\Sigma$. We assume a graph assigns only finitely many edges to $\Plus$ to ensure the finiteness of the data structure on which further computation is done.

@@

We define a graph patch command $\pi = (s,\e) \in \left\{\Plus, \Minus\right\} \times \E$
as a pair of an edge state (excluding $\bot$) and an edge. As a shorthand, we define $+\e$ to be $(\Plus, \e)$ and $-\e$ to be $(\Minus, \e)$.

For $s_1$, $s_2 \in \Sigma$, we define the join operation $s_1 \sqcup s_2$ to be the least upper bound of $s_1$ and $s_2$ with respect to the $\sqsubset$ ordering. Accordingly, $\Plus \sqcup \bot = \Plus$ means that edges can only be created if they do not already exist, and $\Plus \sqcup \Minus = \Minus = \Minus \sqcup \Minus$ means that once an edge is deleted it can never be restored, and hence deletion is idempotent.

We define the semantics of patch commands via the following transition relation
between graphs.
\[
  g \overset{(s,\e)}{\longrightarrow} g\left[ \e \mapsto s \sqcup g(\e) \right]
\]
Applying patch command $\pi = \left(s, \e\right)$ to graph $g$ results in the updated graph $g' = g\left[\e \mapsto s \sqcup g(\e) \right]$, wherein the edge
state associated with edge $\e$ in $g$ becomes the join of $s$ with the state of $\e$ in $g$, and $g'(\e') = g(\e')$ for any $\e' \ne \e$. Essentially, these patch command semantics act as a transition system between graphs, \emph{joining} the new edge state with the corresponding edge state of the graph to which the patch command is being applied. Since a patch command can only change the value of $g$ on one input, applying the patch command preserves the property that $g$ maps only a finite set of edges to $\Plus$.

@@

\begin{lemma}[Join Semilattice]
  \label{lem:Join Semilattice}
  $(\Sigma, \sqsubset)$ with $\sqcup$ forms a join semilattice. That is, $\sqcup$ is associative, commutative, and idempotent. 
\end{lemma}
\begin{proof}
    Follows directly from the total ordering $(\Sigma, \sqsubset)$
\end{proof}

@@

The definition above can be understood operationally as a CmRDT. In particular, the vertices can be understood as forming a Grow-Only Set (G-Set). The edges can be understood as forming a two-phase set (2P-Set), because deletion is permanent~\cite{shapiro2011conflict}.

@@

To help define decomposittion, each vertex in a graph can be classified as either an NP root (if it has no parents), an MP root (if it has multiple parents), a U root (if there is a sequence of unique parents from it to itself, and it has the minimal UID in that sequence), or not a root. These classes are mutually exclusive.

\begin{definition}
    A vertex is an \textit{NP root} in $G$ if $\mathsf{parents}_{G}(v)=\emptyset$
\end{definition}

\begin{definition}
    A vertex is an \textit{MP root} in $G$ if $\abs{\mathsf{parents}_{G}(v)} > 1$
\end{definition}

\begin{definition}
    A vertex is a \textit{U root} in $G$ if $\exists \{v_i\}_{1 \leq i \leq n}$ such that $v_1 = v_n = v$, $\mathsf{parents}_{G}(v_i)=\{v_{i+1}\}$ for all $1 \leq i < n$, and the UID of $v$ is the minimum of the UIDs of $\{v_i\}_{1 \leq i \leq n}$.
\end{definition}

@@

\begin{lemma}
\label{lem:uniq-classification}
    (Unique Classification) For any live subgraph $G\subseteq \E$ and vertex $v\in V$, exactly one of the following holds: 
    \begin{itemize}
        \item $v$ is an NP root in $G$
        \item $v$ is an MP root in $G$
        \item $v$ is an U root in $G$
        \item $v$ is not a root in $G$
    \end{itemize}
\end{lemma}

@@

Formally, the decomposition of a vertex is given by the function $\mathsf{decompVertex}$ a term with the same UID and constructor as the vertex. At each position in the arity of the constructor, we obtain a child term from the function $\mathsf{decompLoc}$, which decomposes a location in the graph to a corresponding child term. Consider the multiplication constructor with id 40 in \autoref{fig:Decomposition example}. It decomposes to a multiplication term with left and right child terms, the result of $\mathsf{decompLoc}$ applied to its left and right locations.

@@

\begin{definition}
\begin{align*}
    \mathsf{decompLoc}_{G}(\ell) &= \begin{cases}
        \ehole &\quad \mathsf{children}_{G}(\ell) = \emptyset \\ \eid{w}{}{\mathsf{decompChild}_G(w, v')} &\quad \mathsf{children}_{G}(\ell) = \SetOf{(w,v')}\\ \conflictHoleTerm{\ell}{\eid{w}{}{\mathsf{decompChild}_G(w, v')} \SuchThat{(w,v') \in \mathsf{children}_{G}(\ell)}} &\quad \abs{\mathsf{children}_{G}(\ell)} > 1 \\
    \end{cases}
\end{align*}
\end{definition}

@@

\begin{definition}
    \mathsf{vertices}(G) = \SetOf{v \in \V \SuchThat{(w, (v,p), v') \in G \land w \in \W \land v' \in \V \land p \in \P}}
\end{definition}

@@

\begin{definition}
    $\mathsf{recomp}(T) = \bigup{\termMV \in T}\mathsf{recompTerm}(\termMV)$
\end{definition}

@@

\begin{theorem}
\label{thm:Recomposability}
    (Recomposability) For all subgraphs $G$, $\mathsf{recomp}(\mathsf{decomp}(G))=G$.
\end{theorem}

@@

Formally, a partitioned grove $\gamma$ is a quadruple $(\subtermMV_r, NP, MP, U)$ where $\subtermMV_r \in ChildTerm$ and $NP, MP, U$ are finite sets of terms. The construction of a partitioned grove from the grove $\mathsf{decomp}(G)$ requires the designation of a \textit{distinguished root} location $\ell_r=(v_r=(u_r,k_r),p_r)$ such that $v_r$ is an NP root in $G$ and $\mathsf{arity}(k_r)=\SetOf{p_r}$. The distinguished child term can be thought of as the decomposition of the distinguished root location. The NP, MP, and U partitions are delineated by the class of the root vertex of each term. The root vertex of a term is like the vertex of a term, except that the root vertex of a reference is the \textit{source} of the corresponding edge, rather than the destination. A term rooted at the distinguished root is excluded from the NP class, since its contents are already identified as the distinguished child term. \autoref{fig:Decomposition example partitioned grove} provides an example of a partitioned grove, where the position $\mathsf{Root}$ of vertex 0 is the distinguished root location.

@@

Edit actions operate on zippered terms~\cite{DBLP:journals/jfp/Huet97,DBLP:conf/popl/OmarVHAH17}. The definitions of zipper terms and child terms are standard, except that the cursor cannot select a child term of the form $\eid{w}{}\termMV$, but is forced to select the term $\termMV$ directly. While inspecting a grove, the user may select an edit action $\alpha$ while the cursor is represented by $\Z{\termMV}$ or $\Z{\subtermMV}$, and this will result in a graph patch $\Q{\pi}$ to be applied directly to the  graph.

@@

The definitions and theorems in \autoref{sub:Convergent Graphs} or \autoref{subsub:Terms}, and \autoref{subsub:Graph Decomposition} have been mechanized in the Agda proof assistant, with the exception that termination is justified separately for some definitions and proofs. In particular, \autoref{thm:Commutativity} and \autoref{thm:Recomposability} are formalized and proven. \autoref{lem:uniq-classification} follows from the more powerful theorem of classification correctness and completeness, which is mechanized and used to prove \autoref{thm:Recomposability}.

>>>>

We now formalize the ideas presented informally above by defining the Grove Calculus. The calculus consists of a commutative patch language on graphs, defined in \autoref{sub:Convergent Graphs}, a decomposition procedure to go from graphs to groves (i.e. sets of terms with conflicts) in \autoref{sub:Groves}, and a user action language that operates on groves and is defined by translation to the patch language in \autoref{sub:User Actions}.

@@

For example, if the language is the simply typed lambda calculus, $\K$ will include constructors for function abstractions ($\textsc{Lambda}_x$), function applications ($\textsc{Ap}$), variables ($\textsc{Var}_x$), and constants ($\textsc{Const}_c$), as well as function type constructors ($\textsc{Arrow}$) and base types ($\textsc{Base}_b$). The arity of each constructor will reflect the child positions of these constructors, e.g. $\mathsf{arity}(\textsc{Ap}) = \{\textsc{Function} , \textsc{Argument}\}$, $\mathsf{arity}(\textsc{Lambda}_x) = \{\textsc{Annotation} , \textsc{Body}\}$, and $\mathsf{arity}(\textsc{Var}_x) = \emptyset$.

@@

A \emph{location} $\ell = (v, p) \in \L$ represents the origin of an edge, where $v=(u,k)$ is drawn from $\V$ and position $p$ is drawn from $\mathsf{arity}(k)$.

An \emph{edge} $\e = (w, \ell, v) \in \E = \W \times \L \times \V$ represents a directed multi-edge 
    identified by $w$, 
    originating from location $\ell$, 
    with destination vertex $v$.

A \emph{graph} $g : \E \rightarrow \Sigma$ is a function from edges to \emph{edge states} $s \in \Sigma = \SetOf{\bot, \Plus, \Minus}$. If $g(\e) = \bot$, then $\e$ has not yet been constructed in the graph. If $g(\e) = \Plus$ then $\e$ is live in the graph. If $g(\e) = \Minus$, then $\e$ has been deleted and cannot be constructed again. The total ordering $\bot \sqsubset \Plus \sqsubset \Minus$ forms a lattice over $\Sigma$. The state of each edge can only progress along this ordering over time, from not yet constructed, to live, to deleted. The \textit{realized} edges in a graph are those assigned to $\Plus$ or $\Minus$. We assume that there are finitely many realized edges in a graph.

@@

We define a graph patch command $\pi = (s,\e) \in \left\{\Plus, \Minus\right\} \times \E$
as a pair of an edge state (excluding $\bot$) and an edge. As a shorthand, we use $+\e$ to denote the construction command $(\Plus, \e)$ and $-\e$ to denote the destruction command $(\Minus, \e)$.

For $s_1$, $s_2 \in \Sigma$, we define the join operation $s_1 \sqcup s_2$ to be the least upper bound of $s_1$ and $s_2$ with respect to the $\sqsubset$ ordering. Accordingly, $\Plus \sqcup \bot = \Plus$ means that an edge can be created if it has never existed, and $\Plus \sqcup \Minus = \Minus$ means that once an edge is deleted it can never be restored.

We define the semantics of patch commands via the following transition relation
between graphs.
\[
  g \overset{(s,\e)}{\longrightarrow} g\left[ \e \mapsto s \sqcup g(\e) \right]
\]
Applying patch command $\pi = \left(s, \e\right)$ to graph $g$ results in the updated graph $g' = g\left[\e \mapsto s \sqcup g(\e) \right]$, wherein the edge
state associated with edge $\e$ in $g$ becomes the join of $s$ with the state of $\e$ in $g$, and $g'(\e') = g(\e')$ for any $\e' \ne \e$. Essentially, these patch command semantics act as a transition system between graphs, \emph{joining} the new edge state with the corresponding edge state of the graph to which the patch command is being applied. Since a patch command can only change the value of $g$ on one input, applying the patch command preserves the property that $g$ maps all but finitely many edges to $\bot$.

@@

\begin{lemma}[Join Semilattice]
  \label{lem:Join Semilattice}
  $(\Sigma, \sqsubset)$ with $\sqcup$ forms a join semilattice. That is, $\sqcup$ is associative, commutative, and idempotent. 
\end{lemma}
\begin{proof}
    % Since $\Sigma$ is finite, 
    Follows directly from the total ordering $(\Sigma, \sqsubset)$. 
\end{proof}

@@

The definition above can be understood operationally as a CmRDT. In particular, the edges can be understood as forming a two-phase set (2P-Set), because deletion is permanent~\cite{shapiro2011conflict}. The set of realized edges forms a grow-only set, as does the set of realized vertices (those included in a realized edge).

@@

To help define decomposition, each vertex in a graph can be classified as either an NP root (if it has no parents), an MP root (if it has multiple parents), a U root (if there is a sequence of unique parents from it to itself, and it has the minimal UID in that sequence), or not a root. These classes are mutually exclusive.

\begin{definition}
    A vertex $v$ is an \textit{NP root} in $G$ if $\mathsf{parents}_{G}(v)=\emptyset$
\end{definition}

\begin{definition}
    A vertex $v$ is an \textit{MP root} in $G$ if $\abs{\mathsf{parents}_{G}(v)} > 1$
\end{definition}

\begin{definition}
    A vertex $v$ is a \textit{U root} in $G$ if $\exists \{v_i\}_{1 \leq i \leq n}$ such that $v_1 = v_n = v$, $\mathsf{parents}_{G}(v_i)=\{v_{i+1}\}$ for all $1 \leq i < n$, and the UID of $v$ is the minimum of the UIDs of $\{v_i\}_{1 \leq i \leq n}$.
\end{definition}

@@

\begin{lemma}[Unique Classification]
\label{lem:uniq-classification}
    For any live subgraph $G\subseteq \E$ and vertex $v\in V$, exactly one of the following holds: 
    \begin{itemize}
        \item $v$ is an NP root in $G$
        \item $v$ is an MP root in $G$
        \item $v$ is an U root in $G$
        \item $v$ is not a root in $G$
    \end{itemize}
\end{lemma}

@@

Formally, the decomposition of a vertex is given by the function $\mathsf{decompVertex}$, which returns a term with the same UID and constructor as the vertex. At each position in the arity of the constructor, we obtain a child term from the function $\mathsf{decompLoc}$, which decomposes a location in the graph to a corresponding child term. Consider the multiplication constructor with id 40 in \autoref{fig:Decomposition example}. It decomposes to a multiplication term with left and right child terms, the result of $\mathsf{decompLoc}$ applied to its left and right locations.

@@

\begin{definition}
\begin{align*}
    \mathsf{decompLoc}_{G}(\ell) &= \begin{cases}
        \ehole &\quad \mathsf{children}_{G}(\ell) = \emptyset \\ \eid{w}{}{\mathsf{decompChild}_G(w, v')} &\quad abs{\mathsf{children}_{G}(\ell)} = 1\\ \conflictHoleTerm{\ell}{\eid{w}{}{\mathsf{decompChild}_G(w, v')} \SuchThat{(w,v') \in \mathsf{children}_{G}(\ell)}} &\quad \abs{\mathsf{children}_{G}(\ell)} > 1 \\
    \end{cases}
\end{align*}
\end{definition}

@@

\begin{definition}
    $\mathsf{vertices}(G) = \SetOf{v \in \V \SuchThat{(w, (v,p), v') \in G \land w \in \W \land v' \in \V \land p \in \P}}$
\end{definition}

@@

\begin{definition}
    $\mathsf{recomp}(T) = \bigcup_{\termMV \in T}\mathsf{recompTerm}(\termMV)$
\end{definition}

@@

\begin{theorem}[Recomposability]
\label{thm:Recomposability}
    For all subgraphs $G$, $\mathsf{recomp}(\mathsf{decomp}(G))=G$.
\end{theorem}

@@

Formally, a partitioned grove $\gamma$ is a quadruple $(\subtermMV_r, NP, MP, U)$ where $\subtermMV_r \in \textit{ChildTerm}$ and $NP, MP, U$ are finite sets of terms. The construction of a partitioned grove from the grove $\mathsf{decomp}(G)$ requires the designation of a \textit{distinguished root} location $\ell_r=(v_r=(u_r,k_r),p_r)$ such that $v_r$ is an NP root in $G$ and $\mathsf{arity}(k_r)=\SetOf{p_r}$. The distinguished child term can be thought of as the decomposition of the distinguished root location. The NP, MP, and U partitions are delineated by the class of the root vertex of each term. The root vertex of a term is like the vertex of a term, except that the root vertex of a reference is the \textit{source} of the corresponding edge, rather than the destination. A term rooted at the distinguished root is excluded from the NP class, since its contents are already identified as the distinguished child term. \autoref{fig:Decomposition example partitioned grove} provides an example of a partitioned grove, where the position $\mathsf{Root}$ of vertex 0 is the distinguished root location.

@@

Edit actions operate on zipper terms~\cite{DBLP:journals/jfp/Huet97,DBLP:conf/popl/OmarVHAH17}. The definitions of zipper terms and child terms are standard, except that the cursor cannot select a child term of the form $\eid{w}{}\termMV$, but is forced to select the term $\termMV$ directly. While inspecting a grove, the user may select an edit action $\alpha$ while the cursor is represented by $\Z{\termMV}$ or $\Z{\subtermMV}$, and this will result in a graph patch $\Q{\pi}$ to be applied directly to the  graph.

@@

The definitions and theorems in \autoref{sub:Convergent Graphs} and \autoref{subsub:Graph Decomposition} have been mechanized in the Agda proof assistant, with the exception that termination is justified separately for some definitions and proofs. In particular, \autoref{thm:Commutativity} and \autoref{thm:Recomposability} are formalized and proven. \autoref{lem:uniq-classification} follows from the more powerful theorem of classification correctness and completeness, which is mechanized and used to prove \autoref{thm:Recomposability}.

### 2024-11-22 sections/grove-by-example.tex
directory restructuring and hiding appendices

This section introduces collaborative structure editing in Grove by example.
\autoref{sub:Program Representation} describes how we use graphs to represent collaborative program sketches. 
\autoref{sub:Single-User Actions} then shows examples of edits being performed by a single user, Alice. 
\autoref{sub:Multi-user Interactions}-\ref{sub:Merge Conflicts} then describes a collaboration between two users, Alice and Bob,
as they edit their own branches of a program and periodically merge in each other's edits, starting with examples without conflicts, then considering the various kinds of conflicts that might arise.

@@

For clarity, we use odd numbers for vertex UIDs and even numbers for edge UIDs. In practice, UIDs would be generated by a mechanism
that effectively ensures that collaborators always generate distinct UIDs, e.g. by generating
universally unique IDs (UUIDs)~\cite{paskin1999toward}.

@@

After converging on \texttt{\hole{} + u * y} in \autoref{fig:DifferentParts:c}, 
Alice changes~\vDifferentPartsAlice{} to~\vNestedPartsAlice{},
producing the edit state in \autoref{fig:NestedParts:a}.
Meanwhile, Bob relocates~\vSimpleTimes{} (bringing along its children)
from the~\texttt{R} position of~\vWrapPlus{} to the~\texttt{L} position. As discussed above, this involves deleting~\eMoveTimes{} and adding~\eNestedPartsBob{}.
Bob's resulting edit state is shown in \autoref{fig:NestedParts:b}.

@@

Grove's support for repositioning creates the possibility for \emph{relocation conflicts}. These occur when a merge causes a vertex to have multiple incoming edges, indicating that it does not have a uniquely determined location (as opposed to local conflicts, which occur when there are multiple outgoing edges at a specified location).

For example, \autoref{fig:MultiParent} shows an example where Alice and Bob relocate a term, \li{w}, to two different locations (the two holes in \autoref{fig:MultiParent:a}). In both cases, the edits are modeled as one edge deletion followed by one edge addition.
Alice deletes~\eNestedPartsAlice{} and adds~\eMultiParentAlice{} in \autoref{fig:MultiParent:b}.
At the same time, Bob deletes~\eNestedPartsAlice{} and adds~\eMultiParentBob{} in \autoref{fig:MultiParent:c}.
Edge deletions is idempotent,
so the fact that Alice and Bob both deleted~\eNestedPartsAlice{} will not lead to a conflict. 
However, both~\eMultiParentAlice{} (added by Alice) and~\eMultiParentBob{} (added by Bob)
point to the same vertex.
Once these patches are merged, the resulting edit state is given in \autoref{fig:MultiParent:d}.
Notice~\vNestedPartsAlice{} has two edges pointing to it. 
When decomposing the graph, we leave a \emph{relocation conflict reference}, $\multiVertex{18}$, at each conflicting location. The conflicted term is separately tracked in the result of decomposition, which, due to conflicts like these, is formally a set of terms with references like these between them. We call this set a \emph{grove}. 
This approach partially addresses the \textbf{relocation conflict problem} from \autoref{sec:Introduction} (we also need to handle unicycles, see below, to fully address the problem).

@@

It is also possible to merge patches such that a \emph{disconnected unicycle} emerges in the graph, even when neither patch disconnects any vertex from the root. \autoref{fig:Disconnection} shows a simple example of when this could occur: Alice relocates~\vMultiCycleTimes{} under~\vSimpleTimes{},
while Bob relocates~\vSimpleTimes{} under~\vMultiCycleTimes{}. 
This results in \autoref{fig:Disconnect:b} and \autoref{fig:Disconnect:c}, respectively.
Merging these causes both vertices to become disconnected, because they were both relocated. The inserted edges form a \emph{unicycle}, meaning a cycle where every vertex has in-degree $1$. In this case, we cannot rely on relocation references to break the cycle. Instead, we break the cycle by arbitrarily but deterministically choosing an edge along the unicycle, e.g. the edge with the smallest UID, and leaving a \emph{unicycle conflict reference} at that location, as shown in \autoref{fig:Disconnect:d}. A user can be notified of this situation when merging and resolve these conflicts again by relocating or deleting terms until the cycle no longer exists in the graph.

>>>>

This section introduces collaborative structure editing in Grove by example.
\autoref{sub:Program Representation} describes how we use graphs to represent collaborative program sketches. 
\autoref{sub:Single-User Actions} then shows examples of edits being performed by a single user, Alice. 
Sections \ref{sub:Multi-user Interactions}--\ref{sub:Merge Conflicts} then describe a collaboration between two users, Alice and Bob,
as they edit their own branches of a program and periodically merge in each other's edits, starting with examples without conflicts, then considering the various kinds of conflicts that might arise.

@@

For clarity, we use even numbers for vertex UIDs and odd numbers for edge UIDs. In practice, UIDs would be generated by a mechanism
that effectively ensures that collaborators always generate distinct UIDs, e.g. by generating
universally unique IDs (UUIDs)~\cite{paskin1999toward}.

@@

After converging on \texttt{\hole{} + u * v} in \autoref{fig:DifferentParts:c}, 
Alice changes~\vDifferentPartsAlice{} to~\vNestedPartsAlice{},
producing the edit state in \autoref{fig:NestedParts:a}.
Meanwhile, Bob relocates~\vSimpleTimes{} (bringing along its children)
from the~\texttt{R} position of~\vWrapPlus{} to the~\texttt{L} position. As discussed above, this involves deleting~\eMoveTimes{} and adding~\eNestedPartsBob{}.
Bob's resulting edit state is shown in \autoref{fig:NestedParts:b}.

@@

Grove's support for code relocation creates the possibility for \emph{relocation conflicts}. These occur when a merge causes a vertex to have multiple incoming edges, indicating that it does not have a uniquely determined location (as opposed to local conflicts, which occur when there are multiple outgoing edges at a specified location).

For example, \autoref{fig:MultiParent} shows an example where Alice and Bob relocate a term, \li{w}, to two different locations (the two holes in \autoref{fig:MultiParent:a}). In both cases, the edits are modeled as one edge deletion followed by one edge addition.
Alice deletes~\eNestedPartsAlice{} and adds~\eMultiParentAlice{} in \autoref{fig:MultiParent:b}.
At the same time, Bob deletes~\eNestedPartsAlice{} and adds~\eMultiParentBob{} in \autoref{fig:MultiParent:c}.
Edge deletion is idempotent,
so the fact that Alice and Bob both deleted~\eNestedPartsAlice{} will not lead to a conflict. 
However, both~\eMultiParentAlice{} (added by Alice) and~\eMultiParentBob{} (added by Bob)
point to the same vertex.
Once these patches are merged, the resulting edit state is given in \autoref{fig:MultiParent:d}.
Notice~\vNestedPartsAlice{} has two edges pointing to it. 
When decomposing the graph, we leave a \emph{relocation conflict reference}, $\multiVertex{18}$, at each conflicting location. The conflicted term is separately tracked in the result of decomposition, which, due to conflicts like these, is formally a set of terms with references between them. We call this set a \emph{grove}. 
This approach partially addresses the \textbf{relocation conflict problem} from \autoref{sec:Introduction} (we also need to handle unicycles, see below, to fully address the problem).

@@

It is also possible to merge patches such that a \emph{disconnected unicycle} emerges in the graph, even when neither patch disconnects any vertex from the root. \autoref{fig:Disconnection} shows a simple example of when this could occur: Alice relocates~\vMultiCycleTimes{} under~\vSimpleTimes{},
while Bob relocates~\vSimpleTimes{} under~\vMultiCycleTimes{}. 
This results in \autoref{fig:Disconnect:b} and \autoref{fig:Disconnect:c}, respectively.
Merging these causes both vertices to become disconnected, because they were both relocated. The inserted edges form a \emph{unicycle}, meaning a cycle where every vertex has in-degree $1$. In this case, we cannot rely on relocation conflict references to break the cycle. Instead, we break the cycle by arbitrarily but deterministically choosing an edge along the unicycle, e.g. the edge with the smallest UID, and leaving a \emph{unicycle conflict reference} at that location, as shown in \autoref{fig:Disconnect:d}. A user can be notified of this situation when merging and resolve these conflicts again by relocating or deleting terms until the cycle no longer exists in the graph.
